---
title: 'Hamilton Christmas Bird Count: Part 2'
subtitle: 'Visualising the Hamilton Christmas Bird Count data'
author: Sharleen
date: '2019-02-23'
slug: hamilton-christmas-bird-count-part-2
summary: Visualising the Hamilton Christmas Bird Count data.
categories: []
tags:
  - data analysis
  - R
  - natural history
  - ornithology
  - ecology
  - birding
  - ggplot2
  - tidyverse
draft: true
image:
  caption: 'Northern Cardinal image credit: [**Public Domain Files**](http://www.publicdomainfiles.com/show_file.php?id=13962429013391)'
  preview_only: false
  
bibliography: part_2.bib
link-citations: true

---

# Note

This is the second part of three for this dataset. [Part 1](https://sharleenw.rbind.io/post/hamilton_cbc_part_1/hamilton-christmas-bird-count-part-1/) contains data downloading and cleaning, Part 2 contains visualisations and Part 3 contains my Shiny app!

# Introduction

A wonderful reference I've been using is the book **Birds of Hamilton and Surrounding Areas** by Robert Curry and the Hamilton Naturalists' Club, particularly the chapter called *Hamilton Christmas Bird Count* by Ian Richards. This book contains a encyclopaedic overview of the history and current state of birds in the Hamilton Study Area.

Let's walk through what I am hoping to learn from this dataset.

1. Compare the first count year's data to the latest year

2. How many species are seen each year

3. Highlight certain species to see if there are any changes over time

To start, we'll load all of the packages we'll be using and set the plot theme:

```{r setup, echo = TRUE, message = FALSE, warning = FALSE}

library(dplyr)
library(janitor)
library(readr)
library(naniar)
library(lubridate)
library(measurements)
library(stringr)
library(ggplot2)
library(here)
library(tidyr)
library(knitr)
library(kableExtra)

theme_set(theme_minimal())

```

```{r read_the_data, include = FALSE, message = FALSE, warning = FALSE}

hamilton_cbc <- read_rds(here("content",
                              "post",
                              "hamilton_cbc_part_3",
                              "hamilton_cbc_shiny",
                              "hamilton_cbc_output.rds"))

```

# Final cleaning touches

While working through this blog post, I realized there were a few more cleaning steps I wanted to do before visualising: 

1. remove counts of `hybrid` birds

2. consolidate the names of some species that have `(` and `/` in them

Let's see how many hybrid species we have and remove them:

``` {r hybrid_removal}

hamilton_cbc %>%
  filter(str_detect(species, "hybrid")) %>%
  distinct(species) %>%
  kable() %>%
  kable_styling(full_width = FALSE, position = "left")

hamilton_cbc <- hamilton_cbc %>%
  filter(!str_detect(species, "hybrid"))

```

Now, onto cleaning the trickier stuff. Sometimes, species have sub-species names or groups, for example:

``` {r junco}

hamilton_cbc %>%
  filter(str_detect(species, "Junco")) %>%
  distinct(species, species_latin) %>%
  kable() %>%
  kable_styling(full_width = FALSE, position = "left")

```

I just want there to be one Dark-eyed Junco species in this dataset, so I am going to consolidate these four sub-species into one species. Even though I know people get way more excited about seeing the Oregon sub-species in Hamilton than the Slate-colored :smile:.

The first step is to only keep the first two words of the `species_latin` variable (this works for all cases in the dataset):

``` {r species_latin_keep_two_words}

hamilton_cbc <- hamilton_cbc %>%
  mutate(species_latin = word(species_latin, start = 1, end = 2))

hamilton_cbc %>%
  filter(str_detect(species, "Junco")) %>%
  distinct(species, species_latin) %>%
  kable() %>%
  kable_styling(full_width = FALSE, position = "left")

```

A bit better! Now let's see who else is in this list:

``` {r multiple_species}

hamilton_cbc %>%
  group_by(species_latin) %>%
  filter(n_distinct(species) > 1) %>%
  ungroup() %>% 
  distinct(species, species_latin) %>%
  kable() %>%
  kable_styling(full_width = FALSE, position = "left")

```

The second step is to sum up the counts for each year across all of these subspecies, and then filter to only keep the first `species` name (which, alphabetically, is the shortest species name):

``` {r subspecies}

hamilton_cbc <- hamilton_cbc %>%
  group_by(year, species_latin) %>%
  mutate(how_many_counted = sum(how_many_counted)) %>%
  arrange(year, species) %>%
  filter(row_number() == 1) %>%
  ungroup()

hamilton_cbc %>%
  filter(str_detect(species, "Junco")) %>%
  distinct(species, species_latin) %>%
  kable() %>%
  kable_styling(full_width = FALSE, position = "left")

```

Perfect! No more sub-species. The last group of species to deal with is where either `species` has a `(` or a `/` or `species_latin` has a `/`:

``` {r name_cleaning}

hamilton_cbc %>%
  group_by(species, species_latin) %>%
  summarise(total_counted = sum(how_many_counted)) %>%
  filter(str_detect(species, "\\(|/") | # The "|" is an "or" within the regex
           str_detect(species_latin, "/")) %>%
  ungroup() %>%
  kable() %>%
  kable_styling(full_width = FALSE, position = "left")

```

I am going to make some executive decisions about what to do with these `species`:

1. Delete species guess: Greater/Lesser Scaup
2. Assume super-rare species were in fact the more common species (please let me know if my bird knowledge is incorrect however!):
    - Bullock's/Baltimore Oriole were Baltimore Orioles
    - Western/Eastern Meadowlark were Eastern Meadowlarks
    - Wilson's/Common Snipe were Common Snipes
    - Spotted/Eastern Towhee (Rufous-sided Towhee) were Eastern Towhees
    - Pacific/Winter Wren were Winter Wrens
3. Remove parentheses on the remaining species for neatness

``` {r removing_parentheses}

hamilton_cbc <- hamilton_cbc %>%
  filter(!(species == "Greater/Lesser Scaup")) %>%
  mutate(species = case_when(species == "Bullock's/Baltimore Oriole" ~ "Baltimore Oriole",
                             species == "Western/Eastern Meadowlark" ~ "Eastern Meadowlark",
                             species == "Wilson's/Common Snipe" ~ "Common Snipe",
                             species == "Spotted/Eastern Towhee (Rufous-sided Towhee)" ~ "Eastern Towhee",
                             species == "Pacific/Winter Wren" ~ "Winter Wren",
                             TRUE ~ species),
         species_latin = case_when(species_latin == "Icterus bullockii/galbula" ~ "Icterus galbula",
                             species_latin == "Sturnella neglecta/magna" ~ "Sturnella magna",
                             species_latin == "Gallinago delicata/gallinago" ~ "Gallinago gallinago",
                             species_latin == "Pipilo maculatus/erythrophthalmus" ~ "Pipilo erythrophthalmus",
                             species_latin == "Troglodytes pacificus/hiemalis" ~ "Troglodytes hiemalis",
                             TRUE ~ species_latin),
         species = case_when(species == "Barn Owl (American)" ~ "Barn Owl",
                             species == "Great Blue Heron (Blue form)" ~ "Great Blue Heron",
                             species == "Palm Warbler (Western)" ~ "Palm Warbler",
                             species == "Rock Pigeon (Feral Pigeon)" ~ "Rock Pigeon",
                             species == "Yellow-rumped Warbler (Myrtle)" ~ "Yellow-rumped Warbler",
                             TRUE ~ species))

# Consolidate the counts between the species whose names were just updated
hamilton_cbc <- hamilton_cbc %>%
  group_by(year, species_latin) %>%
  mutate(how_many_counted = sum(how_many_counted)) %>%
  arrange(year, species) %>%
  filter(row_number() == 1) %>%
  ungroup()

```


Finally, I am just going to recalculate the "per person-hour" variable that depends on `how_many_counted`.

``` {r recalc_person_hours}

hamilton_cbc <- hamilton_cbc %>%
  mutate(how_many_counted_by_hour = as.double(how_many_counted) / total_hours)

```

# Visualising

## Comparing the count's first and latest year

Let's start by looking at the first year anyone went out to do the count, in 1921. In @hsa_birds, Ian Richards mentions that 2 observers went out that year, for 8 hours.

```{r first_year}

hamilton_cbc %>%
  filter(year == min(year),
         how_many_counted != 0) %>%
  select(species, how_many_counted) %>%
  arrange(-how_many_counted, species) %>%
  kable() %>%
  kable_styling(full_width = FALSE, position = "left")

```

Oh wow, they saw a Gray Jay! How many times have Gray Jays been seen on Hamilton CBCs?

```{r gray_jay}

hamilton_cbc %>%
  filter(species == "Gray Jay",
         how_many_counted != 0) %>%
  kable() %>%
  kable_styling(full_width = FALSE, position = "left")

```

That was the only time! At first I thought it might have been a mistake, but @hsa_birds mentions, in the Gray Jay historical records section, that at least two Gray Jays were seen that winter. (This book is so thorough!)

How does the first year's count compare to the most recent year's count?

```{r count_2017}

hamilton_cbc %>%
  filter(year == max(year),
         how_many_counted != 0) %>%
  arrange(-how_many_counted, species) %>%
  head(15) %>%
  kable() %>%
  kable_styling(full_width = FALSE, position = "left")

```

Those species look more familiar! And almost 200 person-hours were put in, amazing!

In 2017, how many species were seen in total?

```{r species_count_2017}

species_count_2017 <- hamilton_cbc %>%
  filter(year == max(year),
         how_many_counted != 0) %>%
  count() %>%
  pull()

```

`r species_count_2017` species, nice! It's not easy to get to 100 species during a CBC. Only four other Ontario CBCs have done it, at least by 2004 (the latest year of CBC data in @hsa_birds).

## Number of species counted each year

Below is a re-creation of a plot found in @hsa_birds. It's a plot of the number of species seen and the number of person-hours put in each year. Person-hours are the total number of hours the counters put in during that count. For example, if 3 people went out for four hours, there were 12 person-hours (4 * 3).

Note: In the course of creating this plot, I believe there was a error in the total hours recorded for 1982, where the total number of hours was only 64, but there was no drop in the number of species counted that year. I think it should have actually been 164 hours, because, in 1981, there were 167 hours, and in 1983, there were 168 hours. So, in the below chunk, I've `mutate`d 1982 to have 164 total hours.

```{r species_and_person_hours}

# Mutating total_hours and how_many_counted_by_hour that depends on it

hamilton_cbc <- hamilton_cbc %>%
  mutate(total_hours = ifelse(year == 1982, 164, total_hours),
         how_many_counted_by_hour = as.double(how_many_counted) / total_hours)


hamilton_cbc %>%
  group_by(year, total_hours) %>% # Bringing total_hours along for the grouping. Adding it won't affect the grouping because its value is constant within the grouping of each year
  filter(how_many_counted != 0) %>%
  summarise(number_of_species_that_year = n()) %>%
  ggplot() +
  geom_line(aes(x = year, y = number_of_species_that_year),
            col = "blue",
            size = 1) +
    geom_line(aes(x = year, y = total_hours),
              col = "orange",
              size = 1) +
  labs(x = "Year",
       y = "Number of species and person-hours")

```

As Ian Richards in @hsa_birds notes, as the number of hours of person-hours increases, so does the number of species counted. But only true up to a point, after which the number of species levels off. This probably means that, as long as there are at least 200 or so person-hours in a count, the Hamilton CBC counters will be finding all or almost all of the species in the Hamilton area on count day.

Here is a plot of the cumulative number of species seen during the count. I think the main conclusion of this plot is that it is getting harder to see new species every year, which makes sense.

```{r cumulative_total_number_of_birds}

hamilton_cbc %>%
  filter(how_many_counted != 0) %>%
  group_by(species) %>%
  filter(year == min(year)) %>% # What's the first year any species was counted?
  ungroup() %>%
  group_by(year) %>%
  summarise(number_of_new_species_that_year = n()) %>%
  mutate(cumulative_new_species = cumsum(number_of_new_species_that_year)) %>%
  ggplot() +
  geom_point(aes(x = year, y = cumulative_new_species)) +
  labs(x = "Year",
       y = "Cumulative number of species counted")

```

## Highlighting specific species

I now want to look at some of the species and how their counts have changed over the course of the count's history. There is a lot to look at!

@hsa_birds mentions that the Hamilton CBC count circle was a different size before 1955, so, for most of the plots, I will only show information starting after 1955. But sometimes I think it's interesting to look at all years of the data, so I might do that too! :smile:

Since the code of these plots will be very similar, I am going to make a function for the plotting, where only the species shown and the start year will vary:

```{r visualising}

plot_species <- function(cbc_df, start_year, species_to_plot) {
  
  cbc_df %>%
    filter(species %in% species_to_plot,
           year >= start_year) %>%
    ggplot() +
    geom_line(aes(
      x = year, 
      y = how_many_counted,
      color = species),
      size = 1) + # Put the size of the line outside of the aes because we're not varying it
    labs(x = "Year",
         y = "Number counted",
         color = "Species") +
    facet_wrap(vars(species),
               scales = "free",
               dir = "v") +
    scale_y_continuous(label = scales::comma) + # Add commas for thousands where necessary
    theme(legend.position = "bottom",
          legend.direction = "vertical")
  
}
```

### Introduced species

Everyone is interested in introduced species. The four I will be plotting are the European Starling, the House Sparrow, the Rock Pigeon, and the Ring-Necked Pheasant.

According to @hsa_birds, Rock Pigeons only started being counted in 1973 and House Sparrows in 1931 (I think due to them being introduced species).

Look at the number of European Starlings! It's so interesting that the numbers peaked in the 1980's and have since seen a sharp decrease. House Sparrows have been pretty consistent. Pigeon numbers have slowly increased, and there was one year, in 2005, with over double the expected number. And the Ring-necked Pheasants have declined to be almost non-existant. @hsa_birds, referencing a 1973 *Waterdown Review* article, says that a change in urbanization pattern, with areas now going directly from intensive farming to sub-divisions, with no scrubby brush period in between (which is where the pheasants liked to live), is the main reason for the decline.

```{r introduced_species}

introduced <- c("European Starling",
                "House Sparrow",
                "Rock Pigeon",
                "Ring-necked Pheasant")

hamilton_cbc %>%
  mutate(how_many_counted = ifelse(species == "Rock Pigeon" &
                                     year < 1973,
                                   NA,
                                   how_many_counted)) %>% # Making Rock Pigeon values pre-1973 to be NA as they were not counted
  plot_species(start_year = 1955,
               species_to_plot = introduced)
```

### Eastern Bluebird

The next species I want to look at is the Eastern Bluebird. I volunteer with a local group that maintains bluebird nest boxes and wanted to see if the Hamilton CBC numbers showed any changes in Eastern Bluebird numbers since the Ontario Eastern Bluebird Society started in 1988.

In @hsa_birds, there is a chapter by Bill Read called *History of the Eastern Bluebird in the Hamilton Study Area*. In that chapter, he states that the reason more bluebirds are overwintering (and thus being counted on the CBC), is mainly due to changing weather patterns. In the first half of the 20th century, bluebirds were common in the spring to fall, but would migrate south in the winter. But since the average winter temperatures have increased since the 1980's, more birds are overwintering. I am not sure why there appear to be such regular increases and decreases since the late 1990's though!

Based on Bill Read's chapter, it sounds as though the nest boxes may not be a major reason in the changing number of bluebirds on the CBC (though they may still be helping bluebird numbers in general).

```{r bluebird}

plot_species(hamilton_cbc,
             start_year = 1955,
             species_to_plot = "Eastern Bluebird")

```

### Wrens

In @hsa_birds, 

I believe the "Pacific/Winter Wren" and "Winter Wren" are in fact the same species (because one of them is counted when the other one is not), so I am going to combine both of those birds into "Winter Wren". Sometimes, birders change what they consider to be a species.

```{r wren}

hamilton_cbc %>%
  filter(str_detect(species, "Wren"),
         species != "House Wren") %>% # I am removing House Wren because of low counts
  mutate(species = ifelse(species == "Pacific/Winter Wren",
                          "Winter Wren",
                          species)) %>%
  group_by(year, species) %>%
  mutate(how_many_counted = sum(how_many_counted)) %>%
  ungroup() %>%
  distinct(year, species, how_many_counted) %>% 
  plot_species(start_year = 1955,
               species_to_plot = .$species)

```



```{r plot}

    # - DDT effect on Bald Eagles and Double-Crested Cormorants
    # - Cardinals: A bird that has been increasing over the past hundred years
    # - A sampling of wrens (Carolina, Marsh and winter), ducks (American Black duck, long-tailed, mallard, ring-necked duck), and gulls (ring-billed and herring)
    # - Chickadees and juncos (slate-coloured and regular)
    # - Cooper's, Sharp-shinned and Red-Tailed Hawks
    # - Snow buntings
    # - Canada goose
    # - Crows, west nile? Are ravens returning to their historical area?
    # - Red-breasted nuthatch and white-breasted nuthatch
    # - Pileated woodpecker (downy, hairy, red-bellied): This was my group's "Bird of the Day" at the most recent CBC


hamilton_cbc %>%
  filter(str_detect(species, "Duck"),
         !(species %in% c("Tufted Duck",
                          "Harlequin Duck",
                          "American Black Duck x Mallard (hybrid)"))) %>% # Removing Tufted and Harlequin ducks due to low counts and removing the hybrid because that's not technically a species (I should probably have removed all hybrids from the whole dataset)
plot_species(1955, .$species)

hamilton_cbc %>%
  filter(str_detect(species_latin, "Larus"),
         !str_detect(species, "hybrid|kumlieni|Thayer")) %>% # Removing gull hybrids or the kumlieni subspecies (?) of the Iceland Gull or the Thayer's Gull (due to the hybrids not being a real species, kumlieni being a subspecies (also with a low count) and Thayer's Gull having a low count)
  plot_species(1955, .$species)

hamilton_cbc %>%
plot_species(start_year = 1955,
             species_to_plot = c("Double-crested Cormorant",
                                 "Bald Eagle")) +
  geom_vline(xintercept = 1973, size = 1) +
  geom_text(x = 1980,
            y = 90,
            label = "1973: DDT was \nbanned in Canada")


# I believe that "Dark-eyed Junco" and "Dark-eyed Junco (Slate-colored)" are different names for the same species (because one of them is counted when the other one is not), so I will combine them into "Dark-eyed Junco". Again, I think this is a case where the official name of this bird has changed over time.

hamilton_cbc %>%
  filter(species %in% c("Dark-eyed Junco",
                        "Dark-eyed Junco (Slate-colored)",
                        "Black-capped Chickadee")) %>%
mutate(species = ifelse(species == "Dark-eyed Junco (Slate-colored)", "Dark-eyed Junco", species)) %>%
  group_by(year, species) %>%
  mutate(how_many_counted = sum(how_many_counted)) %>%
  ungroup() %>%
  distinct(year, species, how_many_counted) %>% 
  plot_species(start_year = 1955,
               species_to_plot = .$species)

plot_species(hamilton_cbc, 1955, c("Cooper's Hawk",
                                   "Sharp-shinned Hawk",
                                   "Red-tailed Hawk"))

hamilton_cbc %>%
  filter(str_detect(species, "Woodpecker"),
         year >= 1955,
         species != "American Three-toed Woodpecker") %>% # Excluding this woodpecker due to low counts
plot_species(1955, .$species)

# American crow declining on pg 306
plot_species(hamilton_cbc, 1955, c("American Crow",
                                   "Common Raven",
                                   "Blue Jay")) +
  geom_vline(xintercept = 2003, size = 1) +
  geom_text(x = 2010,
            y = 4000,
            label = "2003: Height of \nWest Nile virus")

hamilton_cbc %>%
  filter(species %in% c("Red-breasted Nuthatch",
                        "White-breasted Nuthatch")) %>%
  plot_species(1955, .$species)

plot_species(hamilton_cbc, 1955, "Canada Goose")

# Snow buntings had huge peaks in the 1970's
plot_species(hamilton_cbc, 1921, "Snow Bunting")

plot_species(hamilton_cbc, 1921, "Ruffed Grouse")

plot_species(hamilton_cbc, 1921, "Northern Cardinal")

```

This data is only an annual snapshot of the number of birds, and that number depends on several factors, mentioned in @hsa_birds:

  - the number of volunteers
  - the weather of count day
  - the date of the count, which has mostly been on Dec 26 since 1974, but has varied anywhere between Dec 16 and Dec 31

This dataset also doesn't show the populations of species that migrate out of Hamilton in the winter.

In the next blog post, I will be making a Shiny app so that anyone can dig into the data!

And thank you to the CBC! The CBC Data was provided by [National Audubon Society](www.christmasbirdcount.org) and through the generous efforts of [Bird Studies Canada](www.bsc-eoc.org) and countless volunteers across the western hemisphere.

